The Unicode Cookbook \linebreak for Linguists

### 2018-06-08 book/chapters/implementation.tex
deals with paperhivers comments through chapter 5

Searching the Web for ``install r and python'' returns numerous tutorials on how to set up your machine for scientific data analysis. Note that there is no single correct setup for a particular computer or operating system. Both Python and R are available for Windows, Mac, and Unix operating systems from the Python and R project websites. Another option is to use a so-called package manager, i.e.\ a software program that allows the user to manage software packages and their dependencies. On Mac, we use Homebrew,\footnote{\url{https://brew.sh/}} a simple-to-install (via the Terminal App) free and open source package management system. Follow the instructions on the the Homebrew website and then use Homebrew to install R and Python (as well as other software packages such as Git and Jupyter Notebooks).

>>>>

Searching the Web for ``install r and python'' returns numerous tutorials on how to set up your machine for scientific data analysis. Note that there is no single correct setup for a particular computer or operating system. Both Python and R are available for Windows, Mac, and Unix operating systems from the Python and R project websites. Another option is to use a so-called package manager, i.e.\ a software program that allows the user to manage software packages and their dependencies. On Mac, we use Homebrew,\footnote{\url{https://brew.sh/}} a simple-to-install (via the Terminal App) free and open source package management system. Follow the instructions on the Homebrew website and then use Homebrew to install R and Python (as well as other software packages such as Git and Jupyter Notebooks).

### 2018-06-08 book/chapters/ipa_meets_unicode.tex
deals with paperhivers comments through chapter 5

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical glyphs, depending
on the font rendering). For example, a user of a Cyrillic computer keyboard
should ideally not use the <а> \textsc{cyrillic small letter a} at code point
\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
letter a} at code point \uni{0061}, although visually they are mostly
indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
keyboard. Some further problematic homoglyphs related to encoding IPA in the
Unicode Standard are the following.

@@

   \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the real retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
  \item There are two mostly identical-looking Unicode characters for the superscript
        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
        even more problematic is that they have different behavior (see 
        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
        behavior of <ʰ> with other superscript letters, we propose to 
        standardize on the modifier letter at \uni{02B0} (see 
        Section~\ref{pitfall-no-unique-diacritic-ordering}).

@@

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that even in many well-designed fonts the italics of <a> and
<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
the italics of these characters are highly similar <\textit{a}> and
<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
the only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{https://software.sil.org/doulos/}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).} This 
approach was taken by the Language Science Press to distinguish 
between the two different orthographic <a>'s in Chakali in 
\cite{Brindle2017}.\footnote{\url{http://langsci-press.org/catalog/book/74}}

@@

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone (cf.\ \cite{Maddieson1990}). There is an old tradition 
to use diacritics on vowels to mark different tone levels \citep{IPA1949}. Prior to the 1989 Kiel 
convention, IPA-approved tone symbols included diacritics above or below the vowel or syllable, 
e.g.\ high and low tones marked with macrons (<\diaf{0304}>, <\diaf{0331}>), and acute and grave 
accents for high rising tone <\diaf{0301}>, low rising tone <\diaf{0317}>, high falling tone 
<\diaf{0300}> and low falling tone <\diaf{0316}>. These tone symbols, however,  
had failed to catch on (probably) due to aesthetic objections and matters of adequacy for 
transcription \citep[29]{Maddieson1990}.

@@

% IPA1999 pg 19
% "Either of the variant letter shapes [g] and [g] may be used to represent the voiced velar plosive."

Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
question whether this issue is at all apparent to the working linguist, or if
they simply use the \uni{0067} because it is easily keyboarded and thus saves
time, whereas the latter must be cumbersomely inserted as a special symbol in
most software. The International Phonetic Association has taken the stance that
both the keyboard \textsc{latin small letter g} and the \textsc{latin small
letter script g} are valid input characters for the voiced velar plosive \citep[19]{IPA1999}.
Unfortunately, this decision further introduces ambiguity for linguists trying
to adhere to a strict Unicode Standard IPA encoding. For strict IPA encoding we
propose to standardize on the more idiosyncratic \textsc{latin small letter
script g} at \uni{0261}.

@@

Finally, the IPA states that ``diacritics may be placed above a symbol with a
descender''. For example, for marking marking of voiceless pronunciation of
voiced segments the IPA uses the ring diacritic. Originally, the ring should be
placed below the base character, like in <m̥>, using the \textsc{combining ring
below} at \uni{0325}. However, in letters with long descenders the IPA also
allows to put the ring above the base, like in <ŋ̊>, using the \textsc{combining
ring above} at \uni{030A}. Yet, proper font design does not have any problem
with rendering the ring below the base character, like in <ŋ̥>, so for strict
IPA encoding we propose to standardize on the ring below. As a principle, for
strict IPA encoding only one option should be allowed for all diacritics.

@@

In the major revision of the IPA in 1932, affricates were represented by two
consonants <tʃ>, ligatures <ʧ>, or with the tie-bar <t͡ʃ>. In the
1938 revision the tie-bar's semantics were broadened to indicate simultaneous
articulation, as for example in labial velars such as <k͡p>. Thus, the tie-bar is a
convenient diacritic for visually tokenizing input strings into chunks of
phonetically salient groups, including affricates, doubly articulated consonants
or diphthongs.

@@

Further, according to the spirit of the IPA, it would also be possible to
combine more than two base characters into one tie bar, but this is not possible
with Unicode (i.e.~there is no possibility to draw a tie bar over three of four
characters). It is possible to indicate such larger groups by repeating the tie
bar, like for a triphthong <a͡ʊ͡ə> in the English word \textit{hour}. If really
necessary, we consider this possible, even though the rendering will never look
good.

Most importantly though, in comparison to normal Unicode processing, the tie-bar
actually takes a reversed approach to complex graphemes. Basically, the Unicode
principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
sequences in a writing system have to be specified as tailored grapheme
clusters. Only in case the sequence is not a cluster, then this has to be
explicitly indicated. IPA takes a different approach. In IPA by default
different base letters are not connected into larger clusters; only when it is
specified in the string itself (using the tie bar).

@@

Although many combinations of base character with diacritic are treated as
canonical equivalent with precomposed characters, there are a few combinations
in IPA that allow for multiple, apparently identical, encodings that are not
canonical equivalent (see Section~\ref{pitfall-canonical-equivalence} on the
principle of canonical equivalence). For that reason, the following elements
should not be treated as diacritics when encoding IPA in Unicode:
\begin{itemize}
  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0321}> \textsc{combining palatalized hook below} at \uni{0321}
  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0322}> \textsc{combining retroflex hook below} at \uni{0322}
  \item[] <\dia{0335}> \textsc{combining short stroke overlay} at \uni{0335}
  \item[] <\dia{0337}> \textsc{combining short solidus overlay} at \uni{0337}
\end{itemize}

There turn out to be a lot of characters in the IPA that could be conceived as
using any of these elements, like <ɲ>, <ɳ>, <ɨ> or <ø>. However, all such
characters exist as well as precomposed combination in Unicode, and these
precomposed characters should preferably be used. When instead combinations of a base character with
diacritic are used, then these combinations are not canonical equivalent to the
precomposed combinations. This means that any search will not find both at the
same time.

@@

A similar problem arises with the rhotic hook. There are two precomposed
characters in Unicode with a rhotic hook, which are not canonical equivalent 
with a combination of the vowel with a separately encoded hook:
\begin{itemize}
  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
  \item[] <ɝ> \textsc{latin small letter reversed open e with hook} at \uni{025D}
\end{itemize}
All other combinations of vowels with rhotic hooks will have to be made by using
<\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
is no complete set of precomposed characters with rhotic hooks in Unicode. For
that reason we propose to not use the two precomposed characters with hooks
mentioned above, but always use the separate rhotic hook at \uni{02DE} in IPA.\@

@@

Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
\uni{00E7} is canonically equivalent with <c> with <\dia{0327}>
\textsc{combining cedilla} at \uni{0327}, so it will be separated into two
characters by Unicode canonical decomposition, also if such a decomposition is
not intended in the IPA.\@ However, because of the nature of canonical
equivalence (see Section~\ref{pitfall-canonical-equivalence}), these two
encodings are completely identical in any computational treatment, so this
decomposition does not have any practical consequences.

@@

Another pitfall relates to the question of what are diacritics. The problem is that
the meaning of the term diacritics as used by the IPA is not the same as is used
in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
definition combined with the character before them (to form so-called default
grapheme clusters, see Chapter~\ref{the-unicode-approach}). In contrast, Spacing
Modifier Letters are by definition \emph{not} combined into grapheme clusters
with the preceding character, but simply treated as separate letters. In the
context of the IPA, the following IPA-diacritics are actually Spacing Modifier
Letters in the Unicode Standard:

@@

	\item[] Tone letters, including but nut limited to: 
	\begin{itemize} 
	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <˨> \textsc{modifier letter low tone bar} at \uni{02E8}
	\end{itemize}

@@

If intended, then any default combination of Spacing Modifiers
with the preceding character can be specified in orthography profiles (see
Chapter~\ref{orthography-profiles}).

@@

Also related to diacritics is the question of ordering. To our knowledge, the
International Phonetic Association does not specify an ordering for
diacritics that combine with phonetic base symbols; this exercise is left to the
reasoning of the transcriber. However, such marks have to be explicitly ordered
if sequences of them are to be interoperable and compatible computationally. An example is a
labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists
back from using <tʰʷ> instead (with exactly the same intended meaning). However,
from a technical standpoint, these two sequences are different, e.g.~if both
sequences are used in a document, searching for <tʷʰ> will not find any
instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental
nasal can be encoded in various orders, e.g.\ <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.

@@

By Unicode Canonical Combining Classes, overlay diacritics 
like <\dia{0334}> (Combining Class number 1), always come before diacritics
below (Combining Class number 220), which in turn always come before diacritics
above (Combining Class number 230), which in turn come before diacritics over
multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
(Combining Class number 233). We follow this order, but add the other IPA
diacritics (which are not diacritics in the Unicode sense) between diacritics
below and the tie bar. Further, \textit{within} all these classes of diacritics
there is no canonical ordering specified by Unicode, so we propose an explicit
ordering here.

@@

\noindent Then, when a character sequence contains more than one character of the Spacing
Modifier Letters, these will be placed after all combining diacritic marks in the
following order:

@@

The most controversial issue regarding symbols debated at the convention was the
representation for voiceless implosives \citep[62]{Ladefoged1990a}. In
accordance with the principles of the IPA, as outlined in Section
\ref{the-international-phonetic-alphabet}, distinct symbols are favored for
cases of phonological contrast. Further, also convenience of display in the chart
must be taken into account when arguing for or against the inclusion or deletion
of IPA symbols in the IPA chart. Finally, the inclusion or deletion of symbols
should consider the current state of phonetic knowledge of the world's
languages.

\citet{Ladefoged1990a} argued against the inclusion of the symbols < ƥ, ƭ, ƈ, ƙ,
ʠ > for voiceless implosives, noting (i) that they are not contrastive (e.g.\ in
Mayan languages); (ii) that there is no instrumental evidence supporting
voiceless implosives in Africa; and (iii) that the sounds are sufficiently rare
as to not need a whole new row of symbols in the chart. Ladefoged favored
symbolizing the sounds using a voiceless diacritic ring below voiced implosives,
e.g.\ <ɓ̥>. Nevertheless, in the 1989 IPA chart there is indeed a row for
implosives containing voiceless and voiced
pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} But
already in the next revision, in 1993 (with an update in 1996), the voiceless
implosives were dropped. The implosives row from the IPA consonantal chart
disappeared and voiced implosives were given a column in the non-pulmonic
consonants table (which is still reflected in the latest revision to date, IPA
2005).

@@

The point of this pitfall is to highlight that revisions to the IPA will 
continue into the future, albeit they are infrequent. Nevertheless, 
given the Unicode Standard's principle of maintaining backwards compatibility 
(at all costs), transcribers and consumers of IPA cannot rely solely on 
remarks in the Unicode Standard to reflect current standard IPA usage. 
There is the possibility that at a later revision of the IPA, symbols that 
are not currently encoded in the Unicode Standard are added to the IPA -- 
although we think this is unlikely.

>>>>

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical glyphs, depending
on the font rendering). For example, a user of a Cyrillic computer keyboard
should ideally not use the <а> \textsc{cyrillic small letter a} at code point
\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
letter a} at code point \uni{0061}, although visually they are mostly
indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
keyboard. Some further problematic homoglyphs related to encoding IPA in the
Unicode Standard are the following:

@@

   \item The uses of the apostrophe have led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the real retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode character, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
  \item There are two mostly identical-looking Unicode characters for the superscript
        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
        even more problematic is that they have different behavior (see 
        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
        behavior of <ʰ> with other superscript letters, we propose to 
        standardize on the modifier letter at \uni{02B0} (see 
        Section~\ref{pitfall-no-unique-diacritic-ordering}).

@@

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that even in many well-designed fonts the italics of <a> and
<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
the italics of these characters are highly similar, <\textit{a}> and
<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
the only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{https://software.sil.org/doulos/}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).} This 
approach was taken by the Language Science Press to distinguish 
between the two different orthographic <a>'s in Chakali in 
\cite{Brindle2017}.\footnote{\url{http://langsci-press.org/catalog/book/74}}

@@

\noindent The first case in which the IPA allows for different encodings is the question
of how to transcribe tone (cf.\ \cite{Maddieson1990}). There is an old tradition 
to use diacritics on vowels to mark different tone levels \citep{IPA1949}. Prior to the 1989 Kiel 
convention, IPA-approved tone symbols included diacritics above or below the vowel or syllable, 
e.g.\ high and low tones marked with macrons (<\diaf{0304}>, <\diaf{0331}>), and acute and grave 
accents for high rising tone <\diaf{0301}>, low rising tone <\diaf{0317}>, high falling tone 
<\diaf{0300}> and low falling tone <\diaf{0316}>. These tone symbols, however,  
had failed to catch on (probably) due to aesthetic objections and matters of adequacy for 
transcription \citep[29]{Maddieson1990}.

@@

% IPA1999 pg 19
% "Either of the variant letter shapes [g] and [g] may be used to represent the voiced velar plosive."
% https://www.cambridge.org/core/journals/journal-of-the-international-phonetic-association/information/instructions-contributors

Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
question whether this issue is at all apparent to the working linguist, or if
they simply use the \uni{0067} because it is easily keyboarded and thus saves
time, whereas the latter must be cumbersomely inserted as a special symbol in
most software. The International Phonetic Association has taken the stance that
both the keyboard \textsc{latin small letter g} and the \textsc{latin small
letter script g} are valid input characters for the voiced velar plosive 
\citep[19]{IPA1999}.\footnote{Note however that the current instructions for contributors to 
the Journal of the International Phonetic Association requires the use of 
opentail <ɡ> and not looptail <g>.}
Unfortunately, this decision further introduces ambiguity for linguists trying
to adhere to a strict Unicode Standard IPA encoding. For strict IPA encoding we
propose to standardize on the more idiosyncratic \textsc{latin small letter
script g} at \uni{0261}.

@@

Finally, the IPA states that ``diacritics may be placed above a symbol with a
descender''. For example, for marking of voiceless pronunciation of
voiced segments the IPA uses the ring diacritic. Originally, the ring should be
placed below the base character, like in <m̥>, using the \textsc{combining ring
below} at \uni{0325}. However, in letters with long descenders the IPA also
allows to put the ring above the base, like in <ŋ̊>, using the \textsc{combining
ring above} at \uni{030A}. Yet, proper font design does not have any problem
with rendering the ring below the base character, like in <ŋ̥>, so for strict
IPA encoding we propose to standardize on the ring below. As a principle, for
strict IPA encoding only one option should be allowed for all diacritics.

@@

In the major revision of the IPA in 1932, affricates were represented by two
consonants <tʃ>, ligatures <ʧ>, or with the tie bar <t͡ʃ>. In the
1938 revision the tie bar's semantics were broadened to indicate simultaneous
articulation, as for example in labial velars such as <k͡p>. Thus, the tie bar is a
convenient diacritic for visually tokenizing input strings into chunks of
phonetically salient groups, including affricates, doubly articulated consonants
or diphthongs.

@@

Further, according to the spirit of the IPA, it would also be possible to
combine more than two base characters into one tie bar, but this is not possible
with Unicode (i.e.~there is no possibility to draw a tie bar over three or four
characters). It is possible to indicate such larger groups by repeating the tie
bar, like for a triphthong <a͡ʊ͡ə> in the English word \textit{hour}. If really
necessary, we consider this possible, even though the rendering will never look
good.

Most importantly though, in comparison to normal Unicode processing, the tie bar
actually takes a reversed approach to complex graphemes. Basically, the Unicode
principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
sequences in a writing system have to be specified as tailored grapheme
clusters. Only in case the sequence is not a cluster, then this has to be
explicitly indicated. IPA takes a different approach. In IPA by default
different base letters are not connected into larger clusters; only when it is
specified in the string itself (using the tie bar).

@@

Although many combinations of base character with diacritic are treated as
 with precomposed characters, there are a few combinations
in IPA that allow for multiple, apparently identical, encodings that are not
 (see Section~\ref{pitfall-canonical-equivalence} on the
principle of canonical equivalence). For that reason, the following elements
should not be treated as diacritics when encoding IPA in Unicode:
\begin{itemize}
  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0321}> \textsc{combining palatalized hook below} at \uni{0321}
  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0322}> \textsc{combining retroflex hook below} at \uni{0322}
  \item[] <\dia{0335}> \textsc{combining short stroke overlay} at \uni{0335}
  \item[] <\dia{0337}> \textsc{combining short solidus overlay} at \uni{0337}
\end{itemize}

There turn out to be a lot of characters in the IPA that could be conceived as
using any of these elements, like <ɲ>, <ɳ>, <ɨ> or <ø>. However, all such
characters exist as well as precomposed combination in Unicode, and these
precomposed characters should preferably be used. When combinations of a base character with
diacritic are used instead, then these combinations are not  to the
precomposed combinations. This means that any search will not find both at the
same time.

@@

A similar problem arises with the rhotic hook. There are two precomposed
characters in Unicode with a rhotic hook, which are not  
with a combination of the vowel with a separately encoded hook:
\begin{itemize}
  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
  \item[] <ɝ> \textsc{latin small letter reversed open e with hook} at \uni{025D}
\end{itemize}
All other combinations of vowels with rhotic hooks will have to be made by using
<\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
is no complete set of precomposed characters with rhotic hooks in Unicode. For
that reason we propose to not use the two precomposed characters with hooks
mentioned above, but always use the separate rhotic hook at \uni{02DE} in IPA.\@

@@

Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
\uni{00E7} is  with <c> with <\dia{0327}>
\textsc{combining cedilla} at \uni{0327}, so it will be separated into two
characters by Unicode canonical decomposition, also if such a decomposition is
not intended in the IPA.\@ However, because of the nature of canonical
equivalence (see Section~\ref{pitfall-canonical-equivalence}), these two
encodings are completely identical in any computational treatment, so this
decomposition does not have any practical consequences.

@@

Another pitfall relates to the question, what is a diacritic? The problem is that
the meaning of the term \textsc{diacritic} as used by the IPA is not the same as it is used
in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
definition combined with the character before them (to form so-called default
grapheme clusters, see Chapter~\ref{the-unicode-approach}). In contrast, Spacing
Modifier Letters are by definition \emph{not} combined into grapheme clusters
with the preceding character, but simply treated as separate letters. In the
context of the IPA, the following IPA-diacritics are actually Spacing Modifier
Letters in the Unicode Standard:

@@

	\item[] Tone letters, including but not limited to: 
	\begin{itemize} 
	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <˨> \textsc{modifier letter low tone bar} at \uni{02E8}
	\end{itemize}

@@

\noindent If intended, then any default combination of Spacing Modifiers
with the preceding character can be specified in orthography profiles (see
Chapter~\ref{orthography-profiles}).

@@

Also related to diacritics is the question of ordering. To our knowledge, the
International Phonetic Association does not specify an ordering for
diacritics that combine with phonetic base symbols; this exercise is left to the
reasoning of the transcriber. However, such marks have to be explicitly ordered
if sequences of them are to be interoperable and compatible computationally. An example is a
labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists
back from using <tʰʷ> instead (with exactly the same intended meaning). However,
from a technical standpoint, these two sequences are different; if both
sequences are used in a document, searching for <tʷʰ> will not find any
instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental
nasal can be encoded in various orders, e.g.\ <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.

@@

According to Unicode Canonical Combining Classes, overlay diacritics 
like <\dia{0334}> (Combining Class number 1) always come before diacritics
below (Combining Class number 220), which in turn always come before diacritics
above (Combining Class number 230), which in turn come before diacritics over
multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
(Combining Class number 233). We follow this order, but add the other IPA
diacritics (which are not diacritics in the Unicode sense) between diacritics
below and the tie bar. Further, \textit{within} all these classes of diacritics
there is no canonical ordering specified by Unicode, so we propose an explicit
ordering here.

@@

\noindent Then, when a character sequence contains more than one character of the Spacing
Modifier Letters, these will be placed after all combining diacritical marks in the
following order:

@@

The most controversial issue regarding symbols debated at the convention was the
representation for voiceless implosives \citep[62]{Ladefoged1990a}. In
accordance with the principles of the IPA, as outlined in Section
\ref{the-international-phonetic-alphabet}, distinct symbols are favored for
cases of phonological contrast. Further, convenience of display in the chart 
must also be taken into account when arguing for or against the inclusion or deletion
of IPA symbols in the IPA chart. Finally, the inclusion or deletion of symbols
should consider the current state of phonetic knowledge of the world's
languages.

\citet{Ladefoged1990a} argued against the inclusion of the symbols < ƥ, ƭ, ƈ, ƙ,
ʠ > for voiceless implosives, noting (i) that they are not contrastive (e.g.\ in
Mayan languages); (ii) that there is no instrumental evidence supporting
voiceless implosives in Africa; and (iii) that the sounds are sufficiently rare
so as not to need a whole new row of symbols in the chart. Ladefoged favored
symbolizing the sounds using a voiceless diacritic ring below voiced implosives,
e.g.\ <ɓ̥>. Nevertheless, in the 1989 IPA chart there is indeed a row for
implosives containing voiceless and voiced
pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} 
But in the next revision, in 1993 (with an update in 1996), the voiceless
implosives were dropped. The implosives row from the IPA consonantal chart
disappeared and voiced implosives were given a column in the non-pulmonic
consonants table (which is still reflected in the latest revision to date, IPA
2005).

@@

The point of this pitfall is to highlight that revisions to the IPA will 
continue into the future, albeit infrequently. Nevertheless, 
given the Unicode Standard's principle of maintaining backwards compatibility 
(at all costs), transcribers and consumers of IPA cannot rely solely on 
remarks in the Unicode Standard to reflect current standard IPA usage. 
There is the possibility that at a later revision of the IPA, symbols that 
are not currently encoded in the Unicode Standard will be added to the IPA -- 
although we think this is unlikely.
